---
title: "Coursera Reproducible Research-Course Project 1"
author: "Ibraheem Alromaih"
date: "1/9/2019"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, load the necessary libraries

```{r warning=FALSE, message=FALSE}
library(dplyr)
library(mice)
library(timeDate)
library(ggplot2)
```

##1. Code for reading in the dataset and/or processing the data

downlaod the data files
```{r}
file_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url=file_url, "./activity.zip")
```


Unzip the CSV file
```{r}
unzip("./activity.zip")
```

and finally read the CSV file
```{r}
activity <- read.csv("./activity.csv")
```


##2. Histogram of the total number of steps taken each day
Calculate the total daily steps 
```{r}
sum_daily_steps <- activity %>%
  group_by(date) %>%
  summarize(number_of_steps = sum(steps, na.rm = TRUE))
```

plot the histogram of the total daily tem
```{r}
hist(sum_daily_steps$number_of_steps, xlab = "Number of steps per day" ,main = "Histogram of Daily Steps")
```



##3. Mean and median number of steps taken each day
Calculate the mean daily steps
```{r}
mean_daily_steps <- activity %>%
  group_by(date) %>%
  summarize(number_of_steps = mean(steps, na.rm = TRUE))
```

print the results
```{r}
mean_daily_steps
```

Calculate the median daily steps
```{r}
median_daily_steps <- activity %>%
  group_by(date) %>%
  summarize(number_of_steps = median(steps, na.rm = TRUE))
```
print the results
```{r}
median_daily_steps
```

##4. Time series plot of the average number of steps taken
plot the average number of steps
```{r}
plot(mean_daily_steps$number_of_steps ~ mean_daily_steps$date,type="l", xlab="Date", ylab = "Average Number of steps every 5 minutes", main = "Daily Average Steps Every 5 minutes")
```

##5. The 5-minute interval that, on average, contains the maximum number of steps
Find the most active interval
```{r}
most_active_interval <- activity %>%
  group_by(interval) %>%
  summarize(number_of_steps = mean(steps, na.rm = TRUE)) %>%
  filter(number_of_steps == max(number_of_steps))
```
print the results
```{r}
most_active_interval
```

##6. Code to describe and show a strategy for imputing missing data
show the statistic for missing data
```{r}
md.pattern(activity)
```
This shows that 2304 observation has missing steps

impute the missing data
```{r results="hide", warning=FALSE, cache=TRUE}
imputedActivity <- mice(activity,m=5,maxit=50,meth='pmm',seed=500)
```

Add the missing data to the origianl data set
```{r}
completedActivity <- complete(imputedActivity,1)
```

##7. Histogram of the total number of steps taken each day after missing values are imputed
Calculte the sum of steps with the missing data imputed
```{r}
sum_complete_daily_steps <- completedActivity %>%
  group_by(date) %>%
  summarize(number_of_steps = sum(steps, na.rm = TRUE))
```
plot the histogram of the results
```{r}
hist(sum_complete_daily_steps$number_of_steps, xlab = "Number of steps per day" ,main = "Histogram of Daily Steps (missing data imputed)")
```

To further analyze the effect of filling the missing values, the mean and median daily steps are calculated as well:
Calculate the mean daily steps for the data after filling the missing values
```{r}
mean_complete_daily_steps <- completedActivity %>%
  group_by(date) %>%
  summarize(number_of_steps = mean(steps, na.rm = TRUE))
```

print the results
```{r}
mean_complete_daily_steps
```

Calculate the median daily steps
```{r}
median_complete_daily_steps <- completedActivity %>%
  group_by(date) %>%
  summarize(number_of_steps = median(steps, na.rm = TRUE))
```
print the results
```{r}
median_complete_daily_steps
```


##8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
find the average number of steps per 5 minutes interval for weekdays and weekends
```{r}
average_steps_by_interval <- completedActivity %>%
  mutate(isWeekday = ifelse(isWeekday(date, wday=1:5),'Weekdays','Weekends'))  %>%
  group_by(interval,isWeekday) %>%
  summarize(number_of_steps = mean(steps, na.rm = TRUE))
```

Use the data to create a panel plot comparing activity in weekdays and weekends
```{r}
ggplot(data=average_steps_by_interval)+geom_line(aes(x=interval, y=number_of_steps))+facet_grid(isWeekday~.)
```
